---
title: "Lab 1"
author: "Xiangchen Guo"
output: pdf_document
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**What I'm doing:** This lab performs exploratory data analysis using dplyr and ggplot2 to summarize and visualize the data.

```{r}

library(Lahman)
library(dplyr)
data <- Teams %>%
  mutate(
    OBP = (H + BB + HBP) / (AB + BB + HBP + SF),
    SLG = (H + X2B + 2 * X3B + 3 * HR) / AB,
    OPS = OBP + SLG,
    WHIP = (BB + HA) / (IPouts / 3) 
  ) %>%
  select(yearID, teamID, R, OPS, WHIP, FP) %>%
  filter(!is.na(OPS), !is.na(WHIP), !is.na(FP), !is.na(R))
head(data)

#Part 1
model1 <- lm(R ~ OPS + WHIP + FP, data = data)
summary(model1)
par(mfrow=c(2,2)) 
plot(model1)
```
The plots shows up problems that: 
(1) residuals do not appear randomly around 0; 
(2) the residuals deviate significantly from the diagonal line; 
(3) the model fails to account for variability across different scoring levels, potentially due to missing factors or extreme values.
```{r}
#Part 2
data$year_factor <- as.factor(data$yearID)  
league_averages <- data %>%
  group_by(yearID) %>%
  summarise(avgOPS = mean(OPS, na.rm = TRUE),
            avgWHIP = mean(WHIP, na.rm = TRUE),
            avgFP = mean(FP, na.rm = TRUE))
data <- data %>%
  left_join(league_averages, by = "yearID") %>%
  mutate(OPSscale = OPS / avgOPS,
         WHIPscale = WHIP / avgWHIP,
         FPscale = FP / avgFP)
model2 <- lm(log(R) ~ OPSscale + WHIPscale + FPscale + year_factor, data = data)
summary(model2)
par(mfrow = c(2,2))
plot(model2)

```

Model 2 is an improvement over Model 1 in that it incorporates a year variable. It assists in depicting trends over years in the league and smoothes out the Residuals vs Fitted and Scale-Location plots, balancing residuals. With fewer high-leverage points, and according to the Q-Q plot, the data is nearer to being normally distributed, and therefore, the model can accurately model changing scores over years.
```{r}
library(ggplot2)
batters <- Batting %>%
  group_by(playerID) %>%
  summarise(seasons = n_distinct(yearID)) %>%
  filter(seasons >= 10) %>%
  top_n(3, seasons) 

pitchers <- Pitching %>%
  group_by(playerID) %>%
  summarise(seasons = n_distinct(yearID)) %>%
  filter(seasons >= 10) %>%
  top_n(3, seasons)  

batting_stats <- Batting %>%
  filter(playerID %in% batters$playerID) %>%
  mutate(SBpct = SB / (SB + CS),
         OBP = (H + BB + HBP) / (AB + BB + HBP + SF),
         SLG = (H + X2B + 2 * X3B + 3 * HR) / AB,
         OPS = OBP + SLG) %>%
  select(playerID, yearID, G, AB, R, H, X2B, X3B, HR, RBI, SB, CS, SBpct, BB, SO, OBP, SLG, OPS)
batting_stats_clean <- batting_stats %>%
  filter(!is.na(OPS)) 

pitching_stats <- Pitching %>%
  filter(playerID %in% pitchers$playerID) %>%
  mutate(WHIP = (BB + H) / (IPouts / 3),
         SOper9 = (SO / (IPouts / 3)) * 9,
         SOperBB = SO / BB) %>%
  select(playerID, yearID, W, L, IPouts, H, ER, HR, BB, HBP, SO, ERA, WHIP, SOper9, SOperBB)

career_batting <- batting_stats %>%
  group_by(playerID) %>%
  summarise(G = sum(G), AB = sum(AB), R = sum(R), H = sum(H),
            X2B = sum(X2B), X3B = sum(X3B), HR = sum(HR),
            RBI = sum(RBI), SB = sum(SB), CS = sum(CS), BB = sum(BB), 
            SO = sum(SO), OBP = mean(OBP, na.rm = TRUE), 
            SLG = mean(SLG, na.rm = TRUE), OPS = mean(OPS, na.rm = TRUE))

career_pitching <- pitching_stats %>%
  group_by(playerID) %>%
  summarise(W = sum(W), L = sum(L), IPouts = sum(IPouts), 
            H = sum(H), ER = sum(ER), HR = sum(HR), BB = sum(BB),
            HBP = sum(HBP), SO = sum(SO), ERA = mean(ERA, na.rm = TRUE),
            WHIP = mean(WHIP, na.rm = TRUE), 
            SOper9 = mean(SOper9, na.rm = TRUE), 
            SOperBB = mean(SOperBB, na.rm = TRUE))
#Part2
ggplot(batting_stats_clean, aes(x = yearID, y = OPS, color = playerID)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(title = "OPS Career Trajectory for Selected Batters", x = "Year", y = "OPS") 
ggplot(pitching_stats, aes(x = yearID, y = ERA, color = playerID)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(title = "ERA Career Trajectory for Selected Pitchers", x = "Year", y = "ERA") 

```


```{r}
gibson_1968 <- Pitching %>%
  filter(playerID == "gibsobo01" & yearID == 1968) %>%
  select(playerID, yearID, GS, CG, SO, BB, IPouts, H)
 
# (a)
fraction_completed <- gibson_1968$CG / gibson_1968$GS
fraction_completed
# (b)
so_bb_ratio <- gibson_1968$SO / gibson_1968$BB
so_bb_ratio
# (c)
innings_pitched <- gibson_1968$IPouts / 3
innings_pitched
# (d)
whip <- (gibson_1968$H + gibson_1968$BB) / innings_pitched
whip
cat("a) Fraction of games completed by Gibson:", fraction_completed , "\n")
cat("b) Strikeout to walk ratio:", so_bb_ratio, "\n")
cat("c) Innings pitched by Gibson:", innings_pitched, "\n")
cat("d) WHIP:", whip, "\n")
```
 
```{r}
game_data <- data.frame(
  Duration = 139,  
  Attendance = 0,
  VisitorH = 8,  
  VisitorD = 2,  
  VisitorT = 0,
  VisitorHR = 1, 
  VisitorBB = 4, 
  VisitorHBP = 0, 
  VisitorSF = 0,  
  VisitorAB = 32
)
game_duration_hours <- game_data$Duration %/% 60
game_duration_minutes <- game_data$Duration %% 60
extra_base_hits <- game_data$VisitorD + game_data$VisitorT + game_data$VisitorHR
obp <- (game_data$VisitorH + game_data$VisitorBB + game_data$VisitorHBP) /
       (game_data$VisitorAB + game_data$VisitorBB + game_data$VisitorHBP + game_data$VisitorSF)

cat("a) Time:", game_duration_hours, "hours", game_duration_minutes, "minutes\n")
cat("b) Attendance maybe missing or unreported in retrosheet logs.\n")
cat("c) Phillies extra base hits:", extra_base_hits, "\n")
cat("d) Phillies on base percentage:", obp, "\n")

```

